Mortalidad en áreas pequeñas de la región Pampeana (Argentina, 2009-2011)

### 2020-04-06 analysis/Paper_draft_ENG.Rmd
corre hasta la linea 130

---
title: "Mortality Estimates of Departments in Argentina: the case of Pampean Region (2009-2011)"
author: |
  Nicolás Sacco [^ns]; Iván Williams [^iw]; Bernardo L. Queiroz [^bq]

    [^ns]: Cedeplar-UFMG, nsacco@cedeplar.ufmg.br
    [^iw]: Universidad Nacional de Luján, ivanwilliams1985@gmail.com
    [^bq]: Cedeplar-UFMG, lanza@cedeplar.ufmg.br

fontsize: 12pt
foonotesize: 10pt 
date: "Aug, 2019"
output: pdf_document
citation_package: natbib
bibliography: [../bib/refsubmort.bib]
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
# rmarkdown::render('C:/Proyectos/mortalidad_Argentina/analysis/Paper_draft.Rmd', encoding="UTF-8")
```

## **Introduction**

In Latin America and the Caribbean, the demand for epidemiological (and mortality, specifically) estimates of the sub-national heterogeneity is growing, both as a tool for the application of different development plans and for the allocation of resources. The experience in Latin America is led by Brazil, where there is already a sustained methodological development @FreireEtAl2015. In Argentina, experiences were found on estimates of the infant mortality trend (@torcida2008), and the communes of the Autonomous City of Buenos Aires (@Grushka2013). At the same time, more studies are focused in developing demographic tools for measuring individual heterogeneity in a population (@Vaupel_Missov_2013; @vanRaalte_Sasson_Martikainen_2018). In this sense, only considering life table information, there could be multiple layers of inequality analysis just by grouping by administrative levels.

The objective of this work can be enunciated following these progressive steps:  
	+ Estimate the level of mortality of the departments during the period 2009-2011. With the criteria of starting with the population with bigger share to explore methodological approaches, the Pampean region was selected.  
	+ Estimate the heterogeneity between and within departments.  
	+ Test the hypothesis of the negative relationship between internal heterogeneity and the level of mortality at the province level.

In this first work, the step 1 is done ^[All the supplemental material, code and results can be found in https://github.com/nsacco/mortalidad_Argentina. It´s work in progress].

@@

```{r AdjExp, echo = F, fig.align='center', out.width="70%", fig.cap="Exposure Adjustment 2008-2010. Four cases. Source: own based in Census"}
include_graphics('plots/AdjExp.pdf')
```

@@

```{r PF, echo = F, fig.align='center', out.width="70%", fig.cap="Indirect estimates and raw infant mortality. Departments in selected provinces. Source: own based in Census and DEIS"}
include_graphics('plots/ChekPF.pdf')
```

@@

```{r NBI, echo = F, fig.align='center', out.width="70%", fig.cap="Standarized mortality rate and NBI. Departments in selected provinces. Source: own based in Census and DEIS"}
include_graphics('plots/ChekNBI.pdf')
```

@@

```{r cluster, echo = F, fig.align='center', out.width="70%", fig.cap="Regionalization of departments. Source: own based in Census and DEIS"}
include_graphics('plots/cluster.pdf')
```

@@

```{r Ajuste, echo=F, fig.align='center', out.width="70%", fig.cap="Mortality Estimates of Departments. Different methods. Source: own based in Census and DEIS"}
include_graphics('plots/Ajuste.pdf')
```

@@

```{r comparativeMeth, echo = F, fig.align='center', out.width="70%", fig.cap="Estimates of life expectancy at birth with three methods. Source: own based in Census and DEIS"}
include_graphics('plots/CompMethods.pdf')
```

@@

```{r Feos, echo = F, fig.align='center', out.width="70%", fig.cap="Mortality Estimates of Departments with largest differences between methods. Source: own based in Census and DEIS"}
include_graphics('plots/AjusteFeos.pdf')
```

@@

```{r dptosBsAs, echo = F, fig.align='center', out.width=c("70%"), fig.cap="Mortality Estimates of four Departments in Buenos Aires. Source: own based in Census and DEIS"}
include_graphics('plots/dptosBsAs.pdf')
```

>>>>

---
title: "A Methodological Approach Comparison to Estimate Mortality in Small Areas: the Pampean Region in Argentina (2009-2011)"
author: |
  Nicolás Sacco [^ns]; Iván Williams [^iw]; Bernardo L. Queiroz [^bq]

    [^ns]: Penn State, nsacco@psu.edu
    [^iw]: Universidad Nacional de Luján, ivanwilliams1985@gmail.com
    [^bq]: Cedeplar-UFMG, lanza@cedeplar.ufmg.br

fontsize: 12pt
foonotesize: 10pt 
date: "Aug, 2019"
output: pdf_document
citation_package: natbib
bibliography: [../bib/refsubmort.bib]
link-citations: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
# change your dire
load("analysis/Rmarkdown_variables.RData")
# rmarkdown::render('Paper_draft_ESP.Rmd', encoding="UTF-8")
```

# **Abstract**

## **Background**

While the demand for epidemiological estimates of small areas is growing, recent studies show a wide gap and persistent inequalities in life expectancy at birth in Latin America. The lack of geospatial data challenge the application of different methods to study health differentials. Often data in small areas are non-existent or sparce in this region. Spatial patterns are essential to understand individual demographic outcomes related with the characteristics of a place, also as a tool for the application of development plans and for the allocation of resources.

## **Objectives**

The history of estimation in small areas is scarce in Argentina, a clear example of a country with few data sources, but also a very interesting case of the epidemiological transition, often not address by the previous literature. Based on the experience of new applications according to the available auxiliary information, in this paper we apply three different methods to estimate and then compare mortality levels in small areas in the Pampean Region of Argentina, during the period 2009-2011.

## **Methods**

Life expectancy at birht was calculated based on a Bayesian approach, a relational life-table method, smoothing spatial data ‘borrowing strength’ from neighboring areas, and third a indirect demographic approach. Estimates are compared to calculate $q_0$ rates from full death registers and then related to spatial areas, using a simple regionalization analysis.

## **Results**

Results display the potential value of the Bayesian method at the small area-spatial scale. Estimated rates indicate that there is a huge variability in life expectancy at birht withing and between regions, with a spread of up to more than 6 years in the Province of Buenos Aires. Furthermore, we find suggestive evidence that the rates by age present a higher infant mortality but also a higher risk in older adults, in provinces that exibhit more heterogeneity.

---------------------------------------------------------------------------------------------------------------------------------------

@@

```{r AdjExp, echo = F, fig.align='center', out.width="70%", fig.cap="Exposure Adjustment 2008-2010. Four cases. Source: own based in Census"}
include_graphics('analysis/plots/AdjExp.pdf')
```

@@

```{r PF, echo = F, fig.align='center', out.width="70%", fig.cap="Indirect estimates and raw infant mortality. Departments in selected provinces. Source: own based in Census and DEIS"}
include_graphics('analysis/plots/ChekPF.pdf')
```

@@

```{r NBI, echo = F, fig.align='center', out.width="70%", fig.cap="Standarized mortality rate and NBI. Departments in selected provinces. Source: own based in Census and DEIS"}
include_graphics('analysis/plots/ChekNBI.pdf')
```

@@

```{r cluster, echo = F, fig.align='center', out.width="70%", fig.cap="Regionalization of departments. Source: own based in Census and DEIS"}
include_graphics('analysis/plots/cluster.pdf')
```

@@

```{r Ajuste, echo=F, fig.align='center', out.width="70%", fig.cap="Mortality Estimates of Departments. Different methods. Source: own based in Census and DEIS"}
include_graphics('analysis/plots/Ajuste.pdf')
```

@@

```{r comparativeMeth, echo = F, fig.align='center', out.width="70%", fig.cap="Estimates of life expectancy at birth with three methods. Source: own based in Census and DEIS"}
include_graphics('analysis/plots/CompMethods.pdf')
```

@@

```{r Feos, echo = F, fig.align='center', out.width="70%", fig.cap="Mortality Estimates of Departments with largest differences between methods. Source: own based in Census and DEIS"}
include_graphics('analysis/plots/AjusteFeos.pdf')
```

@@

```{r dptosBsAs, echo = F, fig.align='center', out.width=c("70%"), fig.cap="Mortality Estimates of four Departments in Buenos Aires. Source: own based in Census and DEIS"}
include_graphics('analysis/plots/dptosBsAs.pdf')
```

### 2020-04-06 analysis/Paper_draft_ESP.Rmd
corre hasta la linea 130

---
title: "Estimación de Mortalidad por departamentos: el caso de la región Pampeana (2009-2011)"
author: |
  Nicolás Sacco [^ns]; Iván Williams [^iw]; Bernardo L. Queiroz [^bq]

@@

## **Introducción**

@@

@@

```{r AdjExp, echo = F, fig.align='center', out.width="70%", fig.show='hold', fig.cap="Ajuste de exposición 2008-2010. Cuatro casos. Fuente: elaboración propia en base a INDEC (2013)"}
include_graphics('plots/AdjExp.pdf')
```

@@

```{r PF, echo = F, fig.align='center', out.width="70%", fig.cap="Estimaciones indirectas de q(0) y tasa de mortalidad m(0) observada. Departamentos de la región Pampeana (excepto CABA). Fuente: elaboración propia en base a Cesno y estadísticas vitales."}
include_graphics('plots/ChekPF.pdf')
```

@@

```{r cluster, echo = F, fig.align='center', out.width="70%", fig.cap="Regionalización de departamentos. Source: Elaboración propia en base a Censo y Estadísticas Vitales"}
include_graphics('plots/cluster.pdf')
```

@@

```{r comparativeMeth, echo = F, fig.align='center', out.width="70%", fig.cap="Estimaciones de esperanza de vida al nacer según tres metodologías. Source: elaboración propia"}
include_graphics('plots/CompMethods.pdf')
```

@@

```{r Feos, echo = F, fig.align='center', out.width="70%", fig.cap="Estimaciones de mortalidad de los departamentos con mayores diferencias entre métodos. Fuente: elaboración porpia"}
include_graphics('plots/AjusteFeos.pdf')
```

@@

```{r consistAM, echo = F, fig.align='center', fig.height=6, out.width="70%", fig.show='hold', fig.cap="Diferencias relativas en la cantidad de defunciones por edad del área mayor. Fuente: elaboración porpia"}
include_graphics(path = 'plots/ConsistAM.pdf')
```

@@

```{r jerarq, echo = F, fig.align='center', fig.height=6, out.width=c("48%","48%"), fig.show='hold', fig.cap="Esperanza de vida al nacer para las provincias pampeanas (excepto CABA, y Buenos Aires sin La Matanza). Ordenado según la media y con intervalos de confianza. El tamaño de los nombres sigue al tamaño relativo de la población en esa provincia. Fuente: elaboración propia"}
include_graphics(path = c('plots/Buenos_Aires.pdf',
													'plots/Córdoba.pdf',
								 					'plots/Entre_Ríos.pdf',
													'plots/Santa_Fe.pdf',
													'plots/La_Pampa.pdf'))
```

@@

```{r dispersion, echo = F, results='asis', error=F, warning=F}
kable(cbind(provs = provs_names, ineq_df[,-c(1,5)]), digits = 1, caption = 'Resumen de estimaciones por provincia. Esperanza de vida al nacer',
      col.names = c("Provincia", 'Promedio', 'n', 'Varianza', 'Rango')) %>% #'$Ls(IC95)$',
      add_footnote(c("El promedio y varianza no están ponderados", 
      							 "Fuente: elaboración propia"))

@@

<!-- Ampliar un poco-->

###**Un análisis particular: Buenos Aires**

@@

```{r dptosBsAs, echo = F, fig.align='center', fig.height=6, out.width=c("48%","48%"), fig.show='hold', fig.cap="Mortality rates for selected areas in Buenos Aires. Source: own based in Census and DEIS"}
include_graphics("plots/SanIsyMoreno.pdf")
```

@@

>>>>

---
title: "Mortalidad en áreas pequeñas de la región Pampeana (Argentina, 2009-2011)"
author: |
  Nicolás Sacco [^ns]; Iván Williams [^iw]; Bernardo L. Queiroz [^bq]

@@

# **Abstract**

## **Background**

Mientras aumenta la demanada sobre estimaciones epidemiológicas para áreas pequeñas, estudios recientes muestran una persistente brecha de desigualdad en la esperanza de vida al nacer en América Latina. La escasez de datos geoespaciales desafía la aplicación de diferentes métodos para estudiar estos diferenciales de salud. A menudo, los datos en áreas pequeñas o bien no existen o son escasos en la región. Los patrones espaciales son esenciales para comprender los resultados demográficos individuales relacionados con las características de un lugar, también como una herramienta para la aplicación de planes de desarrollo y para la asignación de recursos.

## **Objectives**

La historia de la estimación en áreas pequeñas es escasa en Argentina, un claro ejemplo de un país con pocas fuentes de datos, pero también un caso muy interesante de la transición epidemiológica, que a menudo no se aborda en la literatura previa Basado en la experiencia de nuevas aplicaciones y de acuerdo con la información auxiliar disponible, en este artículo aplicamos tres métodos diferentes para estimar y luego comparar los niveles de mortalidad en áreas pequeñas en la Región Pampeana de Argentina, durante el período 2009-2011.

## **Methods**

Se calculó la esperanza de vida al nacer en base a un enfoque bayesiano, un método de tabla de vida relacional, suavizando datos espaciales de acuerdo a patrones de áreas mayores vecinas, y en tercer lugar se aplicó un enfoque demográfico indirecto. Las estimaciones se compararon para calcular las tasas de $q_0$ a partir de registros de defunción completos y luego se relacionaron con áreas espaciales, utilizando un análisis de regionalización simple.

## **Results**

Los resultados muestran el valor potencial del método bayesiano a pequeña escala espacial-espacial. Las tasas estimadas indican que existe una gran variabilidad en la esperanza de vida al nacer y entre regiones, con una extensión de más de 6 años en la Provincia de Buenos Aires. Además, encontramos evidencia sugestiva de que las tasas por edad presentan una mayor mortalidad infantil, pero también un mayor riesgo en los adultos mayores, en las provincias que muestran más heterogeneidad.

----------------------------------------------------------------------------------------------------------------------------------------------------------------

## **Introduction**

¿Subsisten las disparidades de mortalidad entre las áreas pequeñas? Investigaciones demográficas previas sobre los diferenciales de mortalidad sostienen que los diferenciales geográficos serían gradualmente descartados durante la transición epidemiológica, las políticas de salud pública y las mejoras en la calidad de vida. En la mayoría de los países de ingresos altos y medios, la esperanza de vida al nacer ($q_0$) aumenta a principios del siglo XX, debido al avance del conocimiento y la tecnología médica para combatir las enfermedades infecciosas y las mejoras en las condiciones de vida asociadas con el desarrollo socioeconómico.

Aunque los diferenciales socioeconómicos de mortalidad contribuyeron a mantener la hipotesis de la desigualdad persistente en la muerte durante y después de la transición epidemiológica, algunos autores esperan que los resultados de salud y las disparidades se extingan a medida que se desarrolla el cambio demográfico. Sin embargo, puntos de vista retrospectivos muestran que los diferenciales de mortalidad son persistentes y han crecido con el tiempo en muchos casos de acuerdo a la condición socioeconómica. De todos modos, es útil tener en cuenta si estas disparidades continúan y/o surgen independientemente de  escenarios iniciales auspiciosos.

La explicación teórica más persuasiva de estos temas argumenta que la relación entre mortalidad y posición socioeconómica se ha mantenido a lo largo del tiempo debido al acceso diferencial de clase social a la tecnología de la información y a la salud. Esta hipótesis de causa-efecto condujo a importantes desarrollos de investigación y contribuciones teóricas. Sin embargo, identificar las relaciones basadas en estos patrones y sus resultados demográficos presenta un serio desafío en países donde hay pocas fuentes de datos disponibles. La investigación de la mortalidad implica intrincadas patuas de interacciones sociales, demográficas y ambientales. Por esa razón, la demografía espacial surgió como una perspectiva significativa para responder preguntas demográficas (Matthews, 2013). En ese sentido, los patrones espaciales son esenciales para comprender los resultados demográficos individuales relacionados con las características de un lugar.

Sobre estas cuestiones, aportes metodológicas recientes se han desarrollado para estimaciones de mortalidad en áreas pequeñas. ***Mas sobre aportes recientes para estimacion de mortalidad en areas menores en el mundo.***.

En América Latina y el Caribe, la demanda de estimaciones epidemiológicas (y de mortalidad, específicamente) de la heterogeneidad subnacional está creciendo, tanto como una herramienta para la aplicación de diferentes planes de desarrollo como para la asignación de recursos. La experiencia en América Latina está liderada por Brasil, donde ya existe un desarrollo metodológico sostenido @FreireEtAl2015. En Argentina, se encontraron experiencias sobre estimaciones de la tendencia de mortalidad infantil (@torcida2008) y las comunas de la Ciudad Autónoma de Buenos Aires (@Grushka2013). Al mismo tiempo, otros estudios se centran en el desarrollo de herramientas demográficas para medir la heterogeneidad individual en una población (@Vaupel_Missov_2013; @vanRaalte_Sasson_Martikainen_2018). En este sentido, solo considerando la información de la tabla de vida, podría haber múltiples capas de análisis de desigualdad simplemente agrupando por niveles administrativos.

Argentina representa un claro ejemplo de un país con pocas fuentes de datos, pero también un caso muy interesante de la transición de la mortalidad, que a menudo no se aborda en la literatura previa. Por estas razones, el objetivo de este artículo fue el de elaborar estimaciones de mortalidad que ejemplifiquen la relación entre el condición socioeconómica y salud. Para ello aplicamos tres métodos diferentes para estimar y luego comparar los niveles de mortalidad en áreas pequeñas en la Región Pampeana Argentina, durante el período 2009-2011. Para probar cómo funcionan los métodos elegidos, discriminamos este espacio geográfico ya que es la región más poblada del país.

Utilizamos los datos administrativos del registro de defunciones del Departamento de Estadísticas e Información de Salud del Ministerio de Salud de la Nación (DEIS), unidad oficial del gobierno a cargo de la administración de conteos de defunciones, y las estimaciones de población elaboradas por el Instituto Nacional de Estadísticas y censo (INDEC) agencia gubernamental argentina responsable de la recopilación y el procesamiento de datos estadísticos, como los censos de población. Usando tres enfoques diferentes, primero aplicamos Métodos de Distribución de Muerte (DDM, por su sigla en inglés). Luego combinamos DDM con un enfoque empírico bayesiano, método ya aplicado para el caso de Brasil. El procedimiento de dos pasos nos permite corregir la falta de registro de los recuentos de muertes y minimizar las fluctuaciones aleatorias que pueden ocurrir cuando calculamos los niveles de mortalidad en áreas pequeñas.

# ¿Por qué Argetina?¿Por qué la región Pampeana?

En Argentina, se encontraron experiencias sobre estimaciones de la tendencia de mortalidad infantil (@torcida2008) y las comunas de la Ciudad Autónoma de Buenos Aires (@Grushka2013). Al mismo tiempo, otros estudios se centraron en el desarrollo de herramientas demográficas para medir la heterogeneidad de mortalidad en una población (@Vaupel_Missov_2013; @vanRaalte_Sasson_Martikainen_2018). En este sentido, solo considerando la información de la tabla de vida, podría haber múltiples capas de análisis de desigualdad teniendo en cuenta distintos niveles espaciales.

The lack of geospatial data challenge the application of different methods to study health differentials. Often data in small areas are non-existent or sparce in this region. Spatial patterns are essential to understand individual demographic outcomes related with the characteristics of a place, also as a tool for the application of development plans and for the allocation of resources.

@@

The objective of this work can be enunciated following these progressive steps:  
	+ Estimate the level of mortality of the departments during the period 2009-2011. With the criteria of starting with the population with bigger share to explore methodological approaches, the Pampean region was selected.  
	+ Estimate the heterogeneity between and within departments.  
	+ Test the hypothesis of the negative relationship between internal heterogeneity and the level of mortality at the province level.

In this first work, the step 1 is done ^[All the supplemental material, code and results can be found in https://github.com/nsacco/mortalidad_Argentina. It´s work in progress].

In comparison with other Latin American countries, Argentina’s early socio-economic development, high degree of urbanization, and the expansion of formal education all influenced the reduction of mortality taking place earlier than in most other countries of the region.
This took place due to improvements in living conditions associated with socioeconomic development, rather than the advancement of knowledge and medical technology to fight infectious diseases. Although it took place more quickly and began from higher levels, the drop in mortality in Argentina can, in that sense, be likened to the pattern followed by developed countries with greater distance then from most of the rest of Latin America (Lattes, 1975; Grushka, 2014, p. 169) .

@@

```{r AdjExp, echo = F, fig.align='center', out.width="70%", fig.show='hold', fig.cap="Ajuste de exposición 2008-2010. Cuatro casos. Fuente: elaboración propia en base a INDEC (2013)"}
include_graphics('\analysis/plots/AdjExp.pdf')
```

<!--

@@

```{r PF, echo = F, fig.align='center', out.width="70%", fig.cap="Estimaciones indirectas de q(0) y tasa de mortalidad m(0) observada. Departamentos de la región Pampeana (excepto CABA). Fuente: elaboración propia en base a Cesno y estadísticas vitales."}
include_graphics('analysis/plots/ChekPF.pdf')
```

@@

```{r cluster, echo = F, fig.align='center', out.width="70%", fig.cap="Regionalización de departamentos. Source: Elaboración propia en base a Censo y Estadísticas Vitales"}
include_graphics('analysis/plots/cluster.pdf')
```

@@

```{r comparativeMeth, echo = F, fig.align='center', out.width="70%", fig.cap="Estimaciones de esperanza de vida al nacer según tres metodologías. Source: elaboración propia"}
include_graphics('analysis/plots/CompMethods.pdf')
```

@@

```{r Feos, echo = F, fig.align='center', out.width="70%", fig.cap="Estimaciones de mortalidad de los departamentos con mayores diferencias entre métodos. Fuente: elaboración porpia"}
include_graphics('analysis/plots/AjusteFeos.pdf')
```

@@

```{r consistAM, echo = F, fig.align='center', fig.height=6, out.width="70%", fig.show='hold', fig.cap="Diferencias relativas en la cantidad de defunciones por edad del área mayor. Fuente: elaboración porpia"}
include_graphics(path = 'analysis/plots/ConsistAM.pdf')
```

@@

```{r jerarq, echo = F, fig.align='center', fig.height=6, out.width=c("48%","48%"), fig.show='hold', fig.cap="Esperanza de vida al nacer para las provincias pampeanas (excepto CABA, y Buenos Aires sin La Matanza). Ordenado según la media y con intervalos de confianza. El tamaño de los nombres sigue al tamaño relativo de la población en esa provincia. Fuente: elaboración propia"}
include_graphics(path = c('analysis/plots/Buenos_Aires.pdf',
													'analysis/plots/Córdoba.pdf',
								 					'analysis/plots/Entre_Ríos.pdf',
													'analysis/plots/Santa_Fe.pdf',
													'analysis/plots/La_Pampa.pdf'))
```

@@

<!-- 
```{r dispersion, echo = F, results='asis', error=F, warning=F}
kable(cbind(provs = provs_names, ineq_df[,-c(1,5)]), digits = 1, caption = 'Resumen de estimaciones por provincia. Esperanza de vida al nacer',
      col.names = c("Provincia", 'Promedio', 'n', 'Varianza', 'Rango')) %>% #'$Ls(IC95)$',
      add_footnote(c("El promedio y varianza no están ponderados", 
      							 "Fuente: elaboración propia"))

@@

<!-- Ampliar un poco--> 
<!-- 
###**Un análisis particular: Buenos Aires**

@@

```{r dptosBsAs, echo = F, fig.align='center', fig.height=6, out.width=c("48%","48%"), fig.show='hold', fig.cap="Mortality rates for selected areas in Buenos Aires. Source: own based in Census and DEIS"}
include_graphics("analysis/plots/SanIsyMoreno.pdf")
```

@@

-->
